---
title: "Homework 5 - ML in genomics: Logistic Regression with Penguins Data"
author: "Renan Escalante"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1: Install and Load Necessary Packages

```{r packages, message=FALSE, warning=FALSE}
# Install packages if they are not already installed
# install.packages("tidymodels")
# install.packages("palmerpenguins")

# Load libraries
library(tidymodels)
library(palmerpenguins)
```

## Step 2: Load and Inspect the Data

```{r load_data}
# Load the penguins dataset
data("penguins")

# View the first few rows
head(penguins)
```

## Step 3: Data Preprocessing

### Handle Missing Values

```{r handle_missing}
# Remove rows with missing values
penguins_clean <- penguins %>% 
  na.omit()

# Confirm no missing values remain
sum(is.na(penguins_clean))
```

### Select Relevant Features

```{r select_features}
# Select features and target variable
penguins_df <- penguins_clean %>% 
  select(sex, species, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

# Convert 'sex' to a factor
penguins_df <- penguins_df %>% 
  mutate(sex = as.factor(sex))
```

## Step 4: Split the Data into Training and Testing Sets

```{r split_data}
# Set seed for reproducibility
set.seed(123)

# Split the data (80% training, 20% testing)
penguins_split <- initial_split(penguins_df, prop = 0.8, strata = sex)
penguins_train <- training(penguins_split)
penguins_test <- testing(penguins_split)
```

## Step 5: Specify the Logistic Regression Model

```{r specify_model}
# Specify logistic regression model using glm engine
log_reg_model <- logistic_reg() %>% 
  set_engine("glm")
```

## Step 6: Create a Recipe for Preprocessing

```{r create_recipe}
# Create a recipe for data preprocessing
penguins_recipe <- recipe(sex ~ ., data = penguins_train) %>%
  # Normalize numeric predictors
  step_normalize(all_numeric_predictors()) %>%
  # Create dummy variables for categorical predictors
  step_dummy(all_nominal_predictors())
```

## Step 7: Create a Workflow

```{r create_workflow}
# Combine the model and recipe into a workflow
penguins_workflow <- workflow() %>%
  add_model(log_reg_model) %>%
  add_recipe(penguins_recipe)
```

## Step 8: Fit the Model

```{r fit_model}
# Fit the model to the training data
penguins_fit <- penguins_workflow %>%
  fit(data = penguins_train)
```

## Step 9: Evaluate the Model

### Make Predictions on the Test Set

```{r make_predictions}
# Generate predictions
penguins_predictions <- penguins_fit %>%
  predict(penguins_test) %>%
  bind_cols(penguins_test)
```

### Calculate Performance Metrics

```{r calculate_metrics}
# Calculate accuracy and other metrics
metrics <- penguins_predictions %>%
  metrics(truth = sex, estimate = .pred_class)

# View the metrics
metrics
```

### Confusion Matrix

```{r confusion_matrix}
# Create a confusion matrix
conf_mat <- penguins_predictions %>%
  conf_mat(truth = sex, estimate = .pred_class)

# Visualize the confusion matrix
autoplot(conf_mat, type = "heatmap")
```

## Step 10: Interpret the Results

The performance metrics and confusion matrix will give you insights into how well your logistic regression model is performing in predicting the sex of penguins based on their physical measurements.

* **Accuracy**: The proportion of correct predictions out of all predictions made.
* **Confusion Matrix**: Shows the counts of true positives, true negatives, false positives, and false negatives.

### Optional: Examine Variable Importance

```{r variable_importance}
# Extract the model coefficients
model_coefficients <- tidy(penguins_fit)

# View the coefficients
model_coefficients
```

## Conclusion

You've successfully trained a logistic regression model using `tidymodels` to predict a biological outcome. This example demonstrates how to preprocess data, specify a model, create a workflow, and evaluate the model's performance.

```{r}

```

