---
title: "Homework 6 - ML in genomics: Penguin Classification and Regression with Decision Trees and Random Forests"
author: "Renan Escalante"
date: "2024-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this homework, we'll build both classification and regression models using decision trees and random forests to analyze the Palmer Penguins dataset. We'll predict penguin species (classification) and body mass (regression) using the tidymodels framework.

## Setup

Load the required libraries:

```{r}
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(vip)
```

## Part 1: Classification

### Data Understanding

1. Load the penguins dataset and display the first few rows. What are the features and the target variable for classification?

2. Create a summary of the dataset using `skimr::skim()`. What insights can you gain about the features?

3. Visualize the relationships between features using `GGally::ggpairs()`. What patterns do you notice?

### Data Preparation

4. Remove any rows with missing values from the dataset.

5. Split the data into training and testing sets using `initial_split()`. Use 75% of the data for training and stratify by the species.

6. Create a recipe for data preprocessing for classification. Include the following steps:
   - Normalize all numeric predictors
   - Convert categorical predictors to dummy variables
   - Remove any zero variance predictors

7. Create a validation set using k-fold cross-validation with 5 folds.

### Model Building (Classification)

8. Specify a decision tree model and a random forest model for classification. Set the mode to "classification" for both.

9. Create workflows for both classification models, combining the recipe with each model specification.

10. Fit the classification models to the resamples and collect the performance metrics. Use the following metrics: accuracy, ROC AUC, and Kappa.

### Model Evaluation (Classification)

11. Compare the performance of the decision tree and random forest classification models using the collected metrics. Which model performs better?

12. Create a variable importance plot for the random forest classification model. Which features are most important for predicting penguin species?

13. Fit the best performing classification model to the entire training set and evaluate it on the test set using `last_fit()`. How does the performance on the test set compare to the cross-validated performance?

14. Create a confusion matrix for the classification predictions on the test set. What insights can you gain about the model's performance for each penguin species?

## Part 2: Regression

### Data Preparation (Regression)

15. Create a new recipe for regression, using body_mass_g as the target variable. Include similar preprocessing steps as in the classification recipe.

16. Create a new validation set for regression using k-fold cross-validation with 5 folds.

### Model Building (Regression)

17. Specify a regression tree model using `decision_tree()` and a random forest model using `rand_forest()`. Set the mode to "regression" for both.

18. Create workflows for both regression models, combining the regression recipe with each model specification.

19. Fit the regression models to the resamples and collect the performance metrics. Use the following metrics: RMSE, R-squared, and MAE.

### Model Evaluation (Regression)

20. Compare the performance of the regression tree and random forest regression models using the collected metrics. Which model performs better?

21. Create a variable importance plot for the random forest regression model. Which features are most important for predicting penguin body mass?

22. Fit the best performing regression model to the entire training set and evaluate it on the test set using `last_fit()`. How does the performance on the test set compare to the cross-validated performance?

23. Create a scatter plot of predicted vs. actual body mass values for the test set. What does this plot tell you about the model's performance?

## Conclusion

24. Summarize your findings for both the classification and regression tasks. Discuss the strengths and limitations of using decision trees and random forests for these tasks, and compare their performance in classification vs. regression.